---
title: "The Effects of Pesticide Exposure on Maternal and Child Health"
subtitle: "Final Project BST 260 Fall 2020"
author:
  - Lara Rostomian
  - Sonia Kim
  - Zainab Soetan
output: html_document
---
Overview and Motivation: Provide an overview of the project goals and the motivation for it. Consider that this will be read by people who did not see your project proposal.

Related Work: Anything that inspired you, such as a paper, a web site, or something we discussed in class.

Initial Questions: What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?

Data: Source, scraping method, cleanup, etc.

Exploratory Analysis: What visualizations did you use to look at your data in different ways? What are the different statistical methods you considered? Justify the decisions you made, and show any major changes to your ideas. How did you reach these conclusions?

Final Analysis: What did you learn about the data? How did you answer the questions? How can you justify your answers? Note that 1 type of analysis per team member is required. A Shiny app counts as a type of analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Packages
```{r warning=FALSE}
library(dplyr)
library(tidyverse)
library(pdftools)
library(readr)
library(stringr)
library(ggthemes)
library(shiny)
library(shinyBS)
library(RColorBrewer)
library(shinydashboard)
library(sp)
library(rgeos)
library(rgdal)
library(maptools)
library(leaflet)
library(scales)
library(maps)
library(gridExtra)
library(grid)
```

MCH Data Wrangling *Low BW Only (for Map)
We wanted to create a map to visualize the pattern of low birth weight in California. I initially used CDC WONDER database but it had very limited information, where it grouped rural and small county-level data to "unidentified counties." Thus, I consulted another data source from California Open Data Portal (https://data.ca.gov/dataset/live-births-with-low-and-very-low-birthweight). Following is the data cleaning process. 
```{r}
#Data Wrangling for Year 2014-2018 Data for Map.  
lbwdata<-read.csv("./low-and-very-low-birthweight-by-county-2014-2018 (1).csv", header = TRUE, stringsAsFactors = FALSE)
lbwdata <- lbwdata %>% mutate(County = str_to_title(County))
lbwdata$Events[is.na(lbwdata$Events)] <- 0
lbwdata <- lbwdata %>% group_by(Year, County, Total.Births) %>% summarize(Events = sum(Events)) 
lbwdata <- lbwdata %>% filter(!County == "california") 
lbwdata <- lbwdata %>% mutate(Rate = Events/Total.Births)
``` 

MCH Data Wrangling *Preterm Birth Only (for map)
We also wanted to create another map to visualize the pattern of preterm birth in California. Again, I ran into a similar problem using CDC WONDER database. Thus, I consulted the following database: https://data.chhs.ca.gov/dataset/preterm-and-very-preterm-live-births/resource/cff79e2d-6ecf-4158-9e4f-7078632220ee Following is the cleaning process
```{r}
ptbirthdata<- read.csv("preterm-and-very-preterm-births-by-county-2010-2018-3.csv", header = TRUE, stringsAsFactors = FALSE)
ptbirthdata$Events[is.na(ptbirthdata$Events)] <- 0
ptbirthdata <- ptbirthdata[,-c(7,8)]
ptbirthdata <- ptbirthdata %>% group_by(Year, County, Total.Births) %>% summarize(Events = sum(Events)) 
ptbirthdata <- ptbirthdata %>% filter(!County == "california")#removing the total count
ptbirthdata <- ptbirthdata %>% mutate(rate_pt = Events/Total.Births * 100)

```

Code for Creating the Map - Leaflet Map Using LBW Data (See Shiny App for the Final Result) 
After cleaning the dataset, I then looked at creating a "spatial" map. Following is the wrangling process of generating a map that shapes the map of California, and merging that spatial object with low birth weight data that I wrangled earlier to generate a leaflet map. My main motivation of using a leaflet map was because I wanted to create a map where the user can see which county is which and is able to zoom in and out. Note that there are counties that had NA cases (perhaps for counties that had a very small population)  
```{r}
map <- readOGR(path.expand("cb_2018_us_county_20m.shp"),
               layer = "cb_2018_us_county_20m", stringsAsFactors = FALSE)
Statekey<-read.csv('./STATEFPtoSTATENAME_Key.csv', colClasses=c('character'))
map<-merge(x=map, y=Statekey, by="STATEFP", all=TRUE)
SingleState <- subset(map, map$STATENAME %in% c(
    "California"
))

lbwdata_2016 <- lbwdata %>% filter(Year == "2016") %>% mutate(Rate = Events/Total.Births*100)
spatial_lbw <-sp::merge(x=SingleState, y=lbwdata_2016, by.x="NAME", by.y="County", by=x)

bins <- c(4.0,6.3,7.6,8.1, Inf)
pal <- colorBin(
    palette = "viridis",
    domain = spatial_lbw$Rate, n=7, bins=bins)

leaflet(spatial_lbw, options = leafletOptions(zoomControl = TRUE, zoomLevelFixed = FALSE, dragging=TRUE, minZoom = 5.3, maxZoom = 9)) %>% 
            setView(lat = 36.778259, lng = -119.417931, zoom = 6) %>%
            addPolygons(color = "Black", weight = 1, smoothFactor = 0.5, 
                        opacity = 1.0, fillOpacity = 0.5, layerId = ~NAME,
                        fillColor = ~pal(Rate), 
                        popup = ~as.factor(paste0("<b><font size=\"4\"><center>County: </b>",spatial_lbw$NAME,"</font></center>","<b>% of Low Birth Weight Births: </b>", sprintf("%1.2f%%", spatial_lbw$Rate),"<br/>"))) %>%
            addLegend(pal = pal, values = spatial_lbw$Rate, opacity = 1, title="% Low Birth Weight (Quartiles)")
```

Code for Creating the Map - Leaflet Map Using Preterm Birth Data I Wrangled Earlier (See Shiny App for the Final Result) 
This is a similar spatial map but for pre-term birth. Following is the wrangling process of generating a map that shapes the map of California, and merging that spatial object with pre-term birth data that I wrangled earlier to generate a leaflet map. Note that there are counties that had NA cases (perhaps for counties that had a very small population)  

```{r}
map <- readOGR(path.expand("cb_2018_us_county_20m.shp"),
               layer = "cb_2018_us_county_20m", stringsAsFactors = FALSE)
Statekey<-read.csv('./STATEFPtoSTATENAME_Key.csv', colClasses=c('character'))
map<-merge(x=map, y=Statekey, by="STATEFP", all=TRUE)
SingleState <- subset(map, map$STATENAME %in% c(
    "California"
))

ptbirthdata_2016 <- ptbirthdata %>% filter(Year == "2016") 
spatial_pt <-sp::merge(x=SingleState, y=ptbirthdata_2016, by.x="NAME", by.y="County", by=x)

bin <- c(5.5, 8.2, 9.1, 9.9, Inf)
pal2 <- colorBin(
    palette = "plasma",
    domain = spatial_pt$rate_pt, n=7, bins=bin)

leaflet(spatial_pt, options = leafletOptions(zoomControl = TRUE, zoomLevelFixed = FALSE, dragging=TRUE, minZoom = 5.3, maxZoom = 9)) %>% 
                    setView(lat = 36.778259, lng = -119.417931, zoom = 6) %>%
                    addPolygons(color = "Black", weight = 1, smoothFactor = 0.5, 
                                opacity = 1.0, fillOpacity = 0.5, layerId = ~NAME,
                                fillColor = ~pal2(rate_pt), 
                                popup = ~as.factor(paste0("<b><font size=\"4\"><center>County: </b>",spatial_pt$NAME,"</font></center>","<b>% of Preterm Birth: </b>", sprintf("%1.2f%%", spatial_pt$rate_pt),"<br/>"))) %>% addLegend(pal = pal2, values = spatial_pt$rate_pt, opacity = 1, title="% Preterm Birth (Quartiles)")
            
```

MCH Data Wrangling *New CDC Data (Lara)
```{r}
#Data Wrangling for CDC Data (COMPLETE)
MCH.CDC.Data <- read.delim("NatalityTOTAL.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
MCH.CDC.Data <- MCH.CDC.Data[-c(491:585), ]
MCH.CDC.Data <- MCH.CDC.Data %>% filter(Notes != "Total")
MCH.CDC.Data <- MCH.CDC.Data[ ,-c(1,3,5,7,9)]

MCH.CDC.Data_Race <- read.delim("NatalityRACE.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
MCH.CDC.Data_Race <- MCH.CDC.Data_Race[-c(1794:1931), ]
MCH.CDC.Data_Race <- MCH.CDC.Data_Race[ ,-c(1,3,5,7,9,11)]

MCH.CDC.Data_Care <- read.delim("NatalityCARE.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
MCH.CDC.Data_Care <- MCH.CDC.Data_Care[-c(11063:11487), ]
MCH.CDC.Data_Care <- MCH.CDC.Data_Care[ ,-c(1,3,5,7,9, 11,13)]

#Rename Counties to Match Pesticide Data
MCH.CDC.Data[MCH.CDC.Data$County == "Alameda County, CA", "County"] <-"Alameda"
MCH.CDC.Data[MCH.CDC.Data$County == "Butte County, CA", "County"] <-"Butte"
MCH.CDC.Data[MCH.CDC.Data$County == "Contra Costa County, CA", "County"] <-"Contra Costa"
MCH.CDC.Data[MCH.CDC.Data$County == "El Dorado County, CA", "County"] <-"El Dorado"
MCH.CDC.Data[MCH.CDC.Data$County == "Fresno County, CA", "County"] <-"Fresno"
MCH.CDC.Data[MCH.CDC.Data$County == "Humboldt County, CA", "County"] <-"Humboldt"
MCH.CDC.Data[MCH.CDC.Data$County == "Imperial County, CA", "County"] <-"Imperial"
MCH.CDC.Data[MCH.CDC.Data$County == "Kern County, CA", "County"] <-"Kern"
MCH.CDC.Data[MCH.CDC.Data$County == "Kings County, CA", "County"] <-"Kings"
MCH.CDC.Data[MCH.CDC.Data$County == "Los Angeles County, CA", "County"] <-"Los Angeles"
MCH.CDC.Data[MCH.CDC.Data$County == "Madera County, CA", "County"] <-"Madera"
MCH.CDC.Data[MCH.CDC.Data$County == "Marin County, CA", "County"] <-"Marin"
MCH.CDC.Data[MCH.CDC.Data$County == "Contra Costa County, CA", "County"] <-"Mariposa"
MCH.CDC.Data[MCH.CDC.Data$County == "Merced County, CA", "County"] <-"Merced"
MCH.CDC.Data[MCH.CDC.Data$County == "Monterey County, CA", "County"] <-"Monterey"
MCH.CDC.Data[MCH.CDC.Data$County == "Napa County, CA", "County"] <-"Napa"
MCH.CDC.Data[MCH.CDC.Data$County == "Orange County, CA", "County"] <-"Orange"
MCH.CDC.Data[MCH.CDC.Data$County == "Placer County, CA", "County"] <-"Placer"
MCH.CDC.Data[MCH.CDC.Data$County == "Riverside County, CA", "County"] <-"Riverside"
MCH.CDC.Data[MCH.CDC.Data$County == "Sacramento County, CA", "County"] <-"Sacramento"
MCH.CDC.Data[MCH.CDC.Data$County == "San Bernardino County, CA", "County"] <-"San Bernardino"
MCH.CDC.Data[MCH.CDC.Data$County == "San Diego County, CA", "County"] <-"San Diego"
MCH.CDC.Data[MCH.CDC.Data$County == "San Francisco County, CA", "County"] <-"San Francisco"
MCH.CDC.Data[MCH.CDC.Data$County == "San Joaquin County, CA", "County"] <-"San Joaquin"
MCH.CDC.Data[MCH.CDC.Data$County == "San Luis Obispo County, CA", "County"] <-"San Luis Obispo"
MCH.CDC.Data[MCH.CDC.Data$County == "San Mateo County, CA", "County"] <-"San Mateo"
MCH.CDC.Data[MCH.CDC.Data$County == "Santa Barbara County, CA", "County"] <-"Santa Barbara"
MCH.CDC.Data[MCH.CDC.Data$County == "Santa Clara County, CA", "County"] <-"Canta Clara"
MCH.CDC.Data[MCH.CDC.Data$County == "Santa Cruz County, CA", "County"] <-"Santa Cruz"
MCH.CDC.Data[MCH.CDC.Data$County == "Shasta County, CA", "County"] <-"Shasta"
MCH.CDC.Data[MCH.CDC.Data$County == "Solano County, CA", "County"] <-"Solano"
MCH.CDC.Data[MCH.CDC.Data$County == "Sonoma County, CA", "County"] <-"Sonoma"
MCH.CDC.Data[MCH.CDC.Data$County == "Stanislaus County, CA", "County"] <-"Stanislaus"
MCH.CDC.Data[MCH.CDC.Data$County == "Tulare County, CA", "County"] <-"Tulare"
MCH.CDC.Data[MCH.CDC.Data$County == "Ventura County, CA", "County"] <-"Ventura"
MCH.CDC.Data[MCH.CDC.Data$County == "Yolo County, CA", "County"] <-"Yolo"

MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Alameda County, CA", "County"] <-"Alameda"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Butte County, CA", "County"] <-"Butte"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Contra Costa County, CA", "County"] <-"Contra Costa"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "El Dorado County, CA", "County"] <-"El Dorado"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Fresno County, CA", "County"] <-"Fresno"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Humboldt County, CA", "County"] <-"Humboldt"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Imperial County, CA", "County"] <-"Imperial"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Kern County, CA", "County"] <-"Kern"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Kings County, CA", "County"] <-"Kings"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Los Angeles County, CA", "County"] <-"Los Angeles"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Madera County, CA", "County"] <-"Madera"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Marin County, CA", "County"] <-"Marin"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Contra Costa County, CA", "County"] <-"Mariposa"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Merced County, CA", "County"] <-"Merced"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Monterey County, CA", "County"] <-"Monterey"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Napa County, CA", "County"] <-"Napa"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Orange County, CA", "County"] <-"Orange"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Placer County, CA", "County"] <-"Placer"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Riverside County, CA", "County"] <-"Riverside"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Sacramento County, CA", "County"] <-"Sacramento"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "San Bernardino County, CA", "County"] <-"San Bernardino"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "San Diego County, CA", "County"] <-"San Diego"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "San Francisco County, CA", "County"] <-"San Francisco"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "San Joaquin County, CA", "County"] <-"San Joaquin"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "San Luis Obispo County, CA", "County"] <-"San Luis Obispo"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "San Mateo County, CA", "County"] <-"San Mateo"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Santa Barbara County, CA", "County"] <-"Santa Barbara"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Santa Clara County, CA", "County"] <-"Santa Clara"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Santa Cruz County, CA", "County"] <-"Santa Cruz"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Shasta County, CA", "County"] <-"Shasta"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Solano County, CA", "County"] <-"Solano"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Sonoma County, CA", "County"] <-"Sonoma"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Stanislaus County, CA", "County"] <-"Stanislaus"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Tulare County, CA", "County"] <-"Tulare"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Ventura County, CA", "County"] <-"Ventura"
MCH.CDC.Data_Race[MCH.CDC.Data_Race$County == "Yolo County, CA", "County"] <-"Yolo"

MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Alameda County, CA", "County"] <-"Alameda"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Butte County, CA", "County"] <-"Butte"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Contra Costa County, CA", "County"] <-"Contra Costa"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "El Dorado County, CA", "County"] <-"El Dorado"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Fresno County, CA", "County"] <-"Fresno"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Humboldt County, CA", "County"] <-"Humboldt"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Imperial County, CA", "County"] <-"Imperial"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Kern County, CA", "County"] <-"Kern"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Kings County, CA", "County"] <-"Kings"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Los Angeles County, CA", "County"] <-"Los Angeles"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Madera County, CA", "County"] <-"Madera"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Marin County, CA", "County"] <-"Marin"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Contra Costa County, CA", "County"] <-"Mariposa"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Merced County, CA", "County"] <-"Merced"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Monterey County, CA", "County"] <-"Monterey"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Napa County, CA", "County"] <-"Napa"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Orange County, CA", "County"] <-"Orange"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Placer County, CA", "County"] <-"Placer"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Riverside County, CA", "County"] <-"Riverside"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Sacramento County, CA", "County"] <-"Sacramento"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "San Bernardino County, CA", "County"] <-"San Bernardino"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "San Diego County, CA", "County"] <-"San Diego"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "San Francisco County, CA", "County"] <-"San Francisco"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "San Joaquin County, CA", "County"] <-"San Joaquin"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "San Luis Obispo County, CA", "County"] <-"San Luis Obispo"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "San Mateo County, CA", "County"] <-"San Mateo"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Santa Barbara County, CA", "County"] <-"Santa Barbara"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Santa Clara County, CA", "County"] <-"Santa Clara"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Santa Cruz County, CA", "County"] <-"Santa Cruz"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Shasta County, CA", "County"] <-"Shasta"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Solano County, CA", "County"] <-"Solano"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Sonoma County, CA", "County"] <-"Sonoma"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Stanislaus County, CA", "County"] <-"Stanislaus"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Tulare County, CA", "County"] <-"Tulare"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Ventura County, CA", "County"] <-"Ventura"
MCH.CDC.Data_Care[MCH.CDC.Data_Care$County == "Yolo County, CA", "County"] <-"Yolo"

MCH.CDC.Data_Race<- MCH.CDC.Data_Race %>% rename("Mothers.Race" = "Mother.s.Bridged.Race")
MCH.CDC.Data_Care<- MCH.CDC.Data_Care %>% rename("Mothers.Race" = "Mother.s.Bridged.Race")
``` 

Data Key:
'MCH.CDC.Data' is general data by county 
'MCH.CDC.Data_Race' is stratified by race 
'MCH.CDC.Data_Care' is stratified by race & includes month prenatal care began

Variable Key:
Birth Rate = The number of births divided by total population in the given year(s) [Per 1,000] 


Lara's Plots with CDC Data (Exploratory Analysis)
```{r}
#Fertility Rate GGPLOTS
MCH.CDC.Data %>% 
    ggplot(aes(x = Year, y = County,  fill = Fertility.Rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Fertility Rate", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Fertility Rate by County") + 
    ylab("") + xlab("")

MCH.CDC.Data %>% group_by(County) %>% filter(County == "Fresno") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in Fresno County", subtitle = "2007-2019", color = "County", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data %>% group_by(County) %>% filter(County == "Kern") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in Kern County", subtitle = "2007-2019", color = "County", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data %>% group_by(County) %>% filter(County == "San Joaquin") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in San Joaquin County", subtitle = "2007-2019", color = "County", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data %>% group_by(County) %>% filter(County == "Los Angeles") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in Los Angeles County", subtitle = "2007-2019", color = "County", caption = "Data Source: CDC WONDER Online Database") 


#Grid Plots
p1 <- MCH.CDC.Data %>% group_by(County) %>% filter(County == "Fresno") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in Fresno County", subtitle = "2007-2019", color = "County") + theme(legend.position = "none")

p2 <- MCH.CDC.Data %>% group_by(County) %>% filter(County == "Kern") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in Kern County", subtitle = "2007-2019", color = "County") + theme(legend.position = "none")

p3 <-MCH.CDC.Data %>% group_by(County) %>% filter(County == "San Joaquin") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in San Joaquin County", subtitle = "2007-2019", color = "County") + theme(legend.position = "none")

p4 <-MCH.CDC.Data %>% group_by(County) %>% filter(County == "Los Angeles") %>% ggplot(aes(Year, Fertility.Rate, color = County)) + geom_line() + labs( y="Fertility Rate", title = "Fertility Rate in Los Angeles County", subtitle = "2007-2019", color = "County") + theme(legend.position = "none")

grid.arrange(p1, p2, p3, p4, bottom = "Data Source: CDC WONDER Online Database")
```

```{r}
#Race and Fertility GGPLOTS
MCH.CDC.Data_Race %>% filter(Mothers.Race == "American Indian or Alaska Native") %>%
ggplot(aes(x = Year, y = County,  fill = Fertility.Rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Fertility Rate", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Fertility Rate by County for American Indian or Alaska Native Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Asian or Pacific Islander") %>%
ggplot(aes(x = Year, y = County,  fill = Fertility.Rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Fertility Rate", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Fertility Rate by County for Asian or Pacific Islander Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Black or African American") %>%
ggplot(aes(x = Year, y = County,  fill = Fertility.Rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Fertility Rate", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Fertility Rate by County for Black or African American Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "White") %>%
ggplot(aes(x = Year, y = County,  fill = Fertility.Rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Fertility Rate", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Fertility Rate by County for White Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(County == "Fresno") %>% ggplot(aes(Year, Fertility.Rate, color = Mothers.Race)) + geom_line() + labs( y="Fertility Rate", title = "Race and Fertility Data in Fresno County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")

MCH.CDC.Data_Race %>% filter(County == "Kern") %>% ggplot(aes(Year, Fertility.Rate, color = Mothers.Race)) + geom_line() + labs( y="Fertility Rate", title = "Race and Fertility Data in Kern County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")

MCH.CDC.Data_Race %>% filter(County == "San Bernardino") %>% ggplot(aes(Year, Fertility.Rate, color = Mothers.Race)) + geom_line() + labs( y="Fertility Rate", title = "Race and Fertility Data in San Bernardino County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")

MCH.CDC.Data_Race %>% filter(County == "Los Angeles") %>% ggplot(aes(Year, Fertility.Rate, color = Mothers.Race)) + geom_line() + labs( y="Fertility Rate", title = "Race and Fertility Data in Los Angeles County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")

MCH.CDC.Data_Race %>% filter(County == "San Diego") %>% ggplot(aes(Year, Fertility.Rate, color = Mothers.Race)) + geom_line() + labs( y="Fertility Rate", title = "Race and Fertility Data in San Diego County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")

MCH.CDC.Data_Race %>% filter(County == "Tulare") %>% ggplot(aes(Year, Fertility.Rate, color = Mothers.Race)) + geom_line() + labs( y="Fertility Rate", title = "Race and Fertility Data in Tulare County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")

```

```{r}
#Racial Demographic GGPLOTS
MCH.CDC.Data_Race %>% filter(Mothers.Race == "American Indian or Alaska Native") %>%
    ggplot(aes(x = Year, y = County,  fill = Total.Population)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("American Indian or Alaska Native Population", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Racial Demographics by County") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Asian or Pacific Islander") %>%
    ggplot(aes(x = Year, y = County,  fill = Total.Population)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Asian or Pacific Islander", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Racial Demographics by County") + 
    ylab("") + xlab("")

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Black or African American") %>%
    ggplot(aes(x = Year, y = County,  fill = Total.Population)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Black or African American Population", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Racial Demographics by County") + 
    ylab("") + xlab("")

MCH.CDC.Data_Race %>% filter(Mothers.Race == "White") %>%
    ggplot(aes(x = Year, y = County,  fill = Total.Population)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("White Population", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Racial Demographics by County") + 
    ylab("") + xlab("")

MCH.CDC.Data_Race %>% group_by(County) %>% filter(County == "Fresno") %>% ggplot(aes(Year, Total.Population, color = Mothers.Race)) + geom_line() + scale_y_log10() + labs( y="Total Population (Log 10 Transformation)", title = "Racial Demographics in Fresno County (Transformed Y Axis)", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data_Race %>% group_by(County) %>% filter(County == "Fresno") %>% ggplot(aes(Year, Total.Population, color = Mothers.Race)) + geom_line() + scale_y_log10() + labs( y="Total Population (Log 10 Transformation)", title = "Racial Demographics in Fresno County (Transformed Y Axis)", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data_Race %>% group_by(County) %>% filter(County == "Kern") %>%  ggplot(aes(Year, Total.Population, color = Mothers.Race)) + geom_line()  + scale_y_log10() + labs( y="Total Population (Log 10 Transformation)", title = "Racial Demographics in Kern County (Transformed Y Axis)", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data_Race %>% group_by(County) %>% filter(County == "San Joaquin") %>%  ggplot(aes(Year, Total.Population, color = Mothers.Race)) + geom_line()  + scale_y_log10() + labs( y="Total Population (Log 10 Transformation)", title = "Racial Demographics in San Joaquin County (Transformed Y Axis)", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") 

MCH.CDC.Data_Race %>% group_by(County) %>% filter(County == "Los Angeles") %>%  ggplot(aes(Year, Total.Population, color = Mothers.Race)) + geom_line()  + scale_y_log10() + labs( y="Total Population (Log 10 Transformation)", title = "Racial Demographics in Los Angeles County (Transformed Y Axis)", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  

MCH.CDC.Data_Race %>% group_by(County) %>% filter(County == "Tulare") %>%  ggplot(aes(Year, Total.Population, color = Mothers.Race)) + geom_line()  + scale_y_log10() + labs( y="Total Population (Log 10 Transformation)", title = "Racial Demographics in Tulare County (Transformed Y Axis)", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  
```

```{r}
#Preterm Birth GGPLOTS (With Race)
MCH.CDC.Data %>% 
    ggplot(aes(x = Year, y = County,  fill = Average.LMP.Gestational.Age)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Gestational Age (LMP) in Weeks", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Gestational Age by County") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "American Indian or Alaska Native") %>%
ggplot(aes(x = Year, y = County,  fill = Average.LMP.Gestational.Age)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Gestational Age (LMP) in Weeks", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Gestational Age by County for American Indian or Alaska Native Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Asian or Pacific Islander") %>%
ggplot(aes(x = Year, y = County,  fill = Average.LMP.Gestational.Age)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Gestational Age (LMP) in Weeks", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Gestational Age by County for Asian or Pacific Islander Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Black or African American") %>%
ggplot(aes(x = Year, y = County,  fill = Average.LMP.Gestational.Age)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Gestational Age (LMP) in Weeks", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Gestational Age by County for Black or African American Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "White") %>%
ggplot(aes(x = Year, y = County,  fill = Average.LMP.Gestational.Age)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Gestational Age (LMP) in Weeks", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Gestational Age by County for White Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(County == "Fresno") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in Fresno County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") + geom_hline(yintercept = 37, size =2)

MCH.CDC.Data_Race %>% filter(County == "San Joaquin") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in San Joaquin County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") + geom_hline(yintercept = 37, size =2)

MCH.CDC.Data_Race %>% filter(County == "Kern") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in Kern County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 37, size =2)

MCH.CDC.Data_Race %>% filter(County == "San Bernardino") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in San Bernardino County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 37, size =2)

MCH.CDC.Data_Race %>% filter(County == "Los Angeles") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in Los Angeles County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 37, size =2)

MCH.CDC.Data_Race %>% filter(County == "San Diego") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in San Diego", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 37, size =2)

MCH.CDC.Data_Race %>% filter(County == "Tulare") %>% ggplot(aes(Year, Average.LMP.Gestational.Age, color = Mothers.Race)) + geom_line() + labs( y="Average LMP Gestational Age (Weeks)", title = "Preterm Birth in Tulare", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 37, size =2)

#Preterm Birth Cutoff is 37 Weeks (Horizontal Line)
``` 

```{r}
#Birth weight GGPLOTS (With Race)
MCH.CDC.Data %>% 
    ggplot(aes(x = Year, y = County,  fill = Average.Birth.Weight)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Birth Weight in Grams", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Birth Weight by County") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "American Indian or Alaska Native") %>%
ggplot(aes(x = Year, y = County,  fill = Average.Birth.Weight)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Birth Weight in Grams", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Birth Weight by County for American Indian or Alaska Native Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Asian or Pacific Islander") %>%
ggplot(aes(x = Year, y = County,  fill = Average.Birth.Weight)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Birth Weight in Grams", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Birth Weight by County for Asian or Pacific Islander Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "Black or African American") %>%
ggplot(aes(x = Year, y = County,  fill = Average.Birth.Weight)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Birth Weight Grams", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Birth Weight by County for Black or African American Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(Mothers.Race == "White") %>%
ggplot(aes(x = Year, y = County,  fill = Average.Birth.Weight)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn("Average Birth Weight in Grams", 
                         colors = brewer.pal(9, "Reds")) +
    theme_minimal() +  
    theme(panel.grid = element_blank()) +
    ggtitle("Average Birth Weight by County for White Pop.") + 
    ylab("") + xlab("") 

MCH.CDC.Data_Race %>% filter(County == "Fresno") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in Fresno County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") + geom_hline(yintercept = 2500, size =2)

MCH.CDC.Data_Race %>% filter(County == "San Joaquin") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in San Joaquin County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database") + geom_hline(yintercept = 2500, size =2)

MCH.CDC.Data_Race %>% filter(County == "Kern") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in Kern County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 2500, size =2)

MCH.CDC.Data_Race %>% filter(County == "San Bernardino") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in San Bernardino County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 2500, size =2)

MCH.CDC.Data_Race %>% filter(County == "Los Angeles") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in Los Angeles County", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 2500, size =2)

MCH.CDC.Data_Race %>% filter(County == "San Diego") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in San Diego", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 2500, size =2)

MCH.CDC.Data_Race %>% filter(County == "Tulare") %>% ggplot(aes(Year, Average.Birth.Weight, color = Mothers.Race)) + geom_line() + labs( y="Average Birth Weight (Grams)", title = "Birth Weight Data in Tulare", subtitle = "2007-2019", color = "Mother's Race", caption = "Data Source: CDC WONDER Online Database")  + geom_hline(yintercept = 2500, size =2)

#LBW Cutoff is 2500 Grams (Horizontal Line)

```

Pesticide Data Wrangling
```{r}
county_ranks16 <- read_delim("table1_county_rank_2016.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
repro_lbs16 <- read_delim("table3_reproductive_lbs_2016.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
repro_acre16 <- read_delim("table4_reproductive_acres_2016.txt", "\t", escape_double = FALSE, trim_ws = TRUE)


table1_2016 <- county_ranks16 %>% transmute(county = COUNTY, 
                                     lbs_2015 = LBS_2015, rank_2015 = RANK_2015, 
                                     lbs_2016 = LBS_2016, rank_2016 = RANK_2016)
# column 1 is the county
# columns 2-3 have the previous year data
# columns 4-5 have the current year data

# we only want columns 1-3 for the most up-to-date data for all years before 2016
all_dat <- list(read_csv("table1_2007.csv")[1:3],
                read_csv("table1_2008.csv")[1:3],
                read_csv("table1_2009.csv")[1:3],
                read_csv("table1_2010.csv")[1:3],
                read_csv("table1_2011.csv")[1:3],
                read_csv("table1_2012.csv")[1:3],
                read_csv("table1_2013.csv")[1:3],
                read_csv("table1_2014.csv")[1:3],
                read_csv("table1_2015.csv")[1:3])


table1 <- Reduce(function(x, y) left_join(x, y, by = "county"), all_dat)


long_table1 <- table1 %>% pivot_longer(!county, names_to = 'usage', values_to = "value")
#table1_ranks <- long_table1 %>% filter(str_starts(usage, "rank"))
table1_lbs <- long_table1 %>% filter(str_starts(usage, "lbs"))

long_table2 <- table1_2016 %>% pivot_longer(!county, names_to = 'usage', values_to = "value")
#table1_ranks_1516 <- long_table2 %>% filter(str_starts(usage, "rank"))
table1_lbs_1516<- long_table2 %>% filter(str_starts(usage, "lbs"))

table1_lbs$usage <- as.numeric(gsub("[^[:digit:]]+", "", table1_lbs$usage))
table1_lbs_1516$usage <- as.numeric(gsub("[^[:digit:]]+", "", table1_lbs_1516$usage))
combined_pesticide_use <- table1_lbs %>% full_join(table1_lbs_1516) 
class(combined_pesticide_use$usage) 
combined_pesticide_use <- combined_pesticide_use %>% group_by(usage) 
combined_pesticide_use <- combined_pesticide_use %>% arrange(usage)
```

Sonia's Work with LBW Data:
This is my data wrangling process for low birth weight for the CDC WONDER database. By default, CDC WONDER live birth database only displayed counties that had a county population >100,000. I only looked at low birth rate here and this is for my shiny app bar graph. 
```{r}
#MCH_CDC Data for low birth weight 
#data wrangling mch cdc data
cdc_lowbirthweight <- read.delim("MCH CDC Data.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
cdc_lowbirthweight  <- cdc_lowbirthweight [-c(482:538), ]
cdc_lowbirthweight  <- cdc_lowbirthweight [ ,-c(1, 3, 5, 7)]
MCH.CDC.Data.Total <- read.delim("MCH CDC Data Total.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
MCH.CDC.Data.Total <- MCH.CDC.Data.Total[,-c(1, 3, 5)]
MCH.CDC.Data.Total %>% rename("Total Birth" = "Births")

#I noticed that the data I downloaded did not include total # of births so merging two datasets (one that has total # of birth counts and the other with low birth wegiht +very low birth weight counts)
df1 <- full_join(cdc_lowbirthweight , MCH.CDC.Data.Total, by=c("Year", "County"))
df1<- df1 %>% rename("cases" = "Births.x", "total_births" = "Births.y")

#Note: LBW = Low birth weight + Very low birth weight counts; Total Births = Total # of Birth
col_order <- c("Year", "County", "total_births",
               "cases", "Average.Birth.Weight", "Standard.Deviation.for.Average.Birth.Weight",
               "Average.Age.of.Mother", "Standard.Deviation.for.Average.Age.of.Mother","Average.LMP.Gestational.Age",
               "Standard.Deviation.for.Average.LMP.Gestational.Age")
df2 <- df1[,col_order]
df2[df2$County == "Alameda County, CA", "County"] <-"alameda"
df2[df2$County == "Butte County, CA", "County"] <-"butte"
df2[df2$County == "Contra Costa County, CA", "County"] <-"contra costa"
df2[df2$County == "El Dorado County, CA", "County"] <-"el dorado"
df2[df2$County == "Fresno County, CA", "County"] <-"fresno"
df2[df2$County == "Humboldt County, CA", "County"] <-"humboldt"
df2[df2$County == "Imperial County, CA", "County"] <-"imperial"
df2[df2$County == "Kern County, CA", "County"] <-"kern"
df2[df2$County == "Kings County, CA", "County"] <-"kings"
df2[df2$County == "Los Angeles County, CA", "County"] <-"los angeles"
df2[df2$County == "Madera County, CA", "County"] <-"madera"
df2[df2$County == "Marin County, CA", "County"] <-"marin"
df2[df2$County == "Contra Costa County, CA", "County"] <-"mariposa"
df2[df2$County == "Merced County, CA", "County"] <-"merced"
df2[df2$County == "Monterey County, CA", "County"] <-"monterey"
df2[df2$County == "Napa County, CA", "County"] <-"napa"
df2[df2$County == "Orange County, CA", "County"] <-"orange"
df2[df2$County == "Placer County, CA", "County"] <-"placer"
df2[df2$County == "Riverside County, CA", "County"] <-"riverside"
df2[df2$County == "Sacramento County, CA", "County"] <-"sacramento"
df2[df2$County == "San Bernardino County, CA", "County"] <-"san bernardino"
df2[df2$County == "San Diego County, CA", "County"] <-"san diego"
df2[df2$County == "San Francisco County, CA", "County"] <-"san francisco"
df2[df2$County == "San Joaquin County, CA", "County"] <-"san joaquin"
df2[df2$County == "San Luis Obispo County, CA", "County"] <-"san luis obispo"
df2[df2$County == "San Mateo County, CA", "County"] <-"san mateo"
df2[df2$County == "Santa Barbara County, CA", "County"] <-"santa barbara"
df2[df2$County == "Santa Clara County, CA", "County"] <-"santa clara"
df2[df2$County == "Santa Cruz County, CA", "County"] <-"santa cruz"
df2[df2$County == "Shasta County, CA", "County"] <-"shasta"
df2[df2$County == "Solano County, CA", "County"] <-"solano"
df2[df2$County == "Sonoma County, CA", "County"] <-"sonoma"
df2[df2$County == "Stanislaus County, CA", "County"] <-"stanislaus"
df2[df2$County == "Tulare County, CA", "County"] <-"tulare"
df2[df2$County == "Ventura County, CA", "County"] <-"ventura"
df2[df2$County == "Yolo County, CA", "County"] <-"yolo"
df2 <- df2 %>% filter(!is.na(total_births)) %>% filter(!is.na(cases)) %>% mutate(rate = cases/total_births * 10^2)
df2$County <- df2$County %>% str_to_title()
```

Sonia's CDC WONDER Database Work With Preterm Birth:
This is my data wrangling process for preterm birth for the CDC WONDER database. By default, CDC WONDER live birth database only displayed counties that had a county population >100,000. I only looked at preterm birth here and this is for my shiny app bar graph. 
```{r}
cdc_pretermbirth <- read.delim("Preterm birth.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
cdc_pretermbirth  <- cdc_pretermbirth [-c(422:472), ]
cdc_pretermbirth  <- cdc_pretermbirth [ ,-c(1, 3, 5)]
cdc_pretermbirth <- cdc_pretermbirth %>% rename("Events" = "Births")
MCH.CDC.Data.Total <- read.delim("MCH CDC Data Total.txt",  sep ="\t", dec=".", header = TRUE, stringsAsFactors = FALSE)
MCH.CDC.Data.Total <- MCH.CDC.Data.Total[,-c(1, 3, 5)]
MCH.CDC.Data.Total <- MCH.CDC.Data.Total %>% rename("total_birth" = "Births")

df1_pt <- full_join(cdc_pretermbirth , MCH.CDC.Data.Total, by=c("Year", "County"))

df1_pt[df1_pt$County == "Alameda County, CA", "County"] <-"alameda"
df1_pt[df1_pt$County == "Butte County, CA", "County"] <-"butte"
df1_pt[df1_pt$County == "Contra Costa County, CA", "County"] <-"contra costa"
df1_pt[df1_pt$County == "El Dorado County, CA", "County"] <-"el dorado"
df1_pt[df1_pt$County == "Fresno County, CA", "County"] <-"fresno"
df1_pt[df1_pt$County == "Humboldt County, CA", "County"] <-"humboldt"
df1_pt[df1_pt$County == "Imperial County, CA", "County"] <-"imperial"
df1_pt[df1_pt$County == "Kern County, CA", "County"] <-"kern"
df1_pt[df1_pt$County == "Kings County, CA", "County"] <-"kings"
df1_pt[df1_pt$County == "Los Angeles County, CA", "County"] <-"los angeles"
df1_pt[df1_pt$County == "Madera County, CA", "County"] <-"madera"
df1_pt[df1_pt$County == "Marin County, CA", "County"] <-"marin"
df1_pt[df1_pt$County == "Contra Costa County, CA", "County"] <-"mariposa"
df1_pt[df1_pt$County == "Merced County, CA", "County"] <-"merced"
df1_pt[df1_pt$County == "Monterey County, CA", "County"] <-"monterey"
df1_pt[df1_pt$County == "Napa County, CA", "County"] <-"napa"
df1_pt[df1_pt$County == "Orange County, CA", "County"] <-"orange"
df1_pt[df1_pt$County == "Placer County, CA", "County"] <-"placer"
df1_pt[df1_pt$County == "Riverside County, CA", "County"] <-"riverside"
df1_pt[df1_pt$County == "Sacramento County, CA", "County"] <-"sacramento"
df1_pt[df1_pt$County == "San Bernardino County, CA", "County"] <-"san bernardino"
df1_pt[df1_pt$County == "San Diego County, CA", "County"] <-"san diego"
df1_pt[df1_pt$County == "San Francisco County, CA", "County"] <-"san francisco"
df1_pt[df1_pt$County == "San Joaquin County, CA", "County"] <-"san joaquin"
df1_pt[df1_pt$County == "San Luis Obispo County, CA", "County"] <-"san luis obispo"
df1_pt[df1_pt$County == "San Mateo County, CA", "County"] <-"san mateo"
df1_pt[df1_pt$County == "Santa Barbara County, CA", "County"] <-"santa barbara"
df1_pt[df1_pt$County == "Santa Clara County, CA", "County"] <-"santa clara"
df1_pt[df1_pt$County == "Santa Cruz County, CA", "County"] <-"santa cruz"
df1_pt[df1_pt$County == "Shasta County, CA", "County"] <-"shasta"
df1_pt[df1_pt$County == "Solano County, CA", "County"] <-"solano"
df1_pt[df1_pt$County == "Sonoma County, CA", "County"] <-"sonoma"
df1_pt[df1_pt$County == "Stanislaus County, CA", "County"] <-"stanislaus"
df1_pt[df1_pt$County == "Tulare County, CA", "County"] <-"tulare"
df1_pt[df1_pt$County == "Ventura County, CA", "County"] <-"ventura"
df1_pt[df1_pt$County == "Yolo County, CA", "County"] <-"yolo"
df1_pt <- df1_pt %>% mutate(County = str_to_title(County))
df1_pt <- df1_pt %>% filter(!is.na("total_birth")) %>% filter(!is.na(Events)) %>% mutate(rate = Events/total_birth * 10^2)

```

Sonia's Data Joining of CDC WONDER Live Birth Data (Low Birth Weight and Pre-Term Birth) and Pesticide Data 
I then joined the CDC WONDER data (low birth weight and preterm birth) and Zainab's wrangled pesticide data to come up with a joint data. I then generated bar graphs to visualize the trend across a span of 2007-2016 (Please see shiny app). We noticed that Fresno and Kern county were the two top counties that used the highest amounts of pesticide and found out that San Joaquin Valley is a region that's agriculturally productive. 
```{r}
county_ranks16 <- read_delim("table1_county_rank_2016.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
repro_lbs16 <- read_delim("table3_reproductive_lbs_2016.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
repro_acre16 <- read_delim("table4_reproductive_acres_2016.txt", "\t", escape_double = FALSE, trim_ws = TRUE)

table1_2016 <- county_ranks16 %>% transmute(county = COUNTY, 
                                            lbs_2015 = LBS_2015, rank_2015 = RANK_2015, 
                                            lbs_2016 = LBS_2016, rank_2016 = RANK_2016)

all_dat <- list(read_csv("table1_2007.csv")[1:3],
                read_csv("table1_2008.csv")[1:3],
                read_csv("table1_2009.csv")[1:3],
                read_csv("table1_2010.csv")[1:3],
                read_csv("table1_2011.csv")[1:3],
                read_csv("table1_2012.csv")[1:3],
                read_csv("table1_2013.csv")[1:3],
                read_csv("table1_2014.csv")[1:3],
                read_csv("table1_2015.csv")[1:3])


table1 <- Reduce(function(x, y) left_join(x, y, by = "county"), all_dat)

long_table1 <- table1 %>% pivot_longer(!county, names_to = 'usage', values_to = "value")
table1_ranks <- long_table1 %>% filter(str_starts(usage, "rank"))
table1_lbs <- long_table1 %>% filter(str_starts(usage, "lbs"))
table1_lbs$usage <- as.numeric(gsub("[^[:digit:]]+", "", table1_lbs$usage))

long_table2 <- table1_2016 %>% pivot_longer(!county, names_to = 'usage', values_to = "value")
table1_ranks_1516 <- long_table2 %>% filter(str_starts(usage, "rank"))
table1_lbs_1516<- long_table2 %>% filter(str_starts(usage, "lbs"))


table1_lbs_1516$usage <- as.numeric(gsub("[^[:digit:]]+", "", table1_lbs_1516$usage))
combined_pesticide_use <- table1_lbs %>% full_join(table1_lbs_1516) 
combined_pesticide_use <- combined_pesticide_use %>% group_by(usage) 
combined_pesticide_use <- combined_pesticide_use %>% arrange(usage)

averagebw <-df2 %>% select("County", "Year", "rate")
pesticide_averagebw_join <- averagebw %>% inner_join(combined_pesticide_use, by = c("County" = "county", "Year" = "usage")) 

averagept <-df1_pt %>% select("County", "Year", "rate")
pesticide_averagept_join <- averagept %>% inner_join(combined_pesticide_use, by = c("County" = "county", "Year" = "usage")) 

#bar graph of low birth weight
pesticide_averagebw_join %>% ggplot(aes(County, rate)) + geom_col() + ylab("Low Birth Weight Rate (%)") +xlab("") +
                theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1/2)) 

#bar graph of pesticide 
pesticide_averagept_join %>% ggplot(aes(County, value)) + geom_col() + ylab("Pesticide Use (Pounds)") +xlab("") + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1/2))

```

Code for Creating the Map - Leaflet Map Using Preterm Birth Data I Wrangled Earlier (See Shiny App for the Final Result) 
This is a similar spatial map but for pesticide us. Following is the wrangling process of generating a map that shapes the map of California, and merging that spatial object with pesticide use that I wrangled earlier to generate a leaflet map. 
```{r}
averagept_df <- combined_pesticide_use %>% filter(usage == "2016")

map <- readOGR(path.expand("cb_2018_us_county_20m.shp"),
               layer = "cb_2018_us_county_20m", stringsAsFactors = FALSE)
Statekey<-read.csv('./STATEFPtoSTATENAME_Key.csv', colClasses=c('character'))
map<-merge(x=map, y=Statekey, by="STATEFP", all=TRUE)
SingleState <- subset(map, map$STATENAME %in% c(
    "California"
))

spatial_pesticide <-sp::merge(x=SingleState, y=averagept_df, by.x="NAME", by.y="county", by=x)

binpes <- c(200, 100145, 1131454, 3345277, Inf)
pal3 <- colorBin(
    palette = "magma",
    domain = spatial_pesticide$value, n=7, bins=binpes)

leaflet(spatial_pesticide, options = leafletOptions(zoomControl = TRUE, zoomLevelFixed = FALSE, dragging=TRUE, minZoom = 5.3, maxZoom = 9)) %>% 
                setView(lat = 36.778259, lng = -119.417931, zoom = 6) %>%
                addPolygons(color = "Black", weight = 1, smoothFactor = 0.5, 
                            opacity = 1.0, fillOpacity = 0.5, layerId = ~NAME,
                            fillColor = ~pal3(value), 
                            popup = ~as.factor(paste0("<b><font size=\"4\"><center>County: </b>",spatial_pesticide$NAME,"</font></center>","Amounts of Pesticides used </b>", sprintf("%1.2f", spatial_pesticide$value),"<br/>"))) %>%
                addLegend(pal = pal3, values = spatial_pesticide$value, opacity = 1, title="Amounts of Pesticide Used (Pounds)")
        
    
```
```